---
title: "ragged experiment demos"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{ragged experiment demos}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r getpks,echo=FALSE, results="hide"}
suppressPackageStartupMessages({
  library(curatedTCGAData)
  library(TCGAutils)
  library(BiocParallel)
  library(survival)
})
```

# Enumerate mutations by symbol in BRCA, couple to overall survival

```{r cache=TRUE}
register(MulticoreParam(4))
suppressMessages({
bmut = curatedTCGAData("BRCA", "Mutation", dry.run=FALSE, version="2.0.1")
})

bmut.srv <- TCGAprimaryTumors(bmut)
sampleTables(bmut.srv)
anyReplicated(bmut.srv)
replicates <- Filter(length,
    lapply(replicated(bmut.srv)[[1]], function(x) colnames(bmut.srv[[1]])[x])
)

sum(sapply(replicated(bmut.srv), any))
## 9
## remove replicated observations
bmut.surv <- bmut.srv[, list("BRCA_Mutation-20160128" = !colnames(bmut.srv)[[1]] %in% unlist(sapply(replicates, tail, -1)))]
stopifnot(!anyReplicated(bmut.surv))

## See mutation classifications
table(mcols(bmut.surv[[1]])$Variant_Classification)

# set up survival time
colData(bmut.surv) <- colData(bmut)[!is.na(bmut$OS.Time), ]
osurv <- Surv(bmut.surv$OS.Time/365.25, bmut.surv$OS.event)

system.time(
  mutsyms <- as(lapply(as(bmut.surv[[1L]], "GRangesList"), function(x) x$Hugo_Symbol), "CharacterList")
)

system.time(
  lapply(mutsyms, function(x) sum(duplicated(x)))
)
```

```{r lksurv}
plot(survfit(osurv~1))
```

The display, based on 793 observations with non-missing
TCGA OS.Time, is consistent with the display at the [GDC Data Portal](https://portal.gdc.cancer.gov/exploration?filters=%7B%22op%22%3A%22and%22%2C%22content%22%3A%5B%7B%22content%22%3A%7B%22field%22%3A%22cases.diagnoses.tissue_or_organ_of_origin%22%2C%22value%22%3A%5B%22axillary%20tail%20of%20breast%22%2C%22breast%2C%20nos%22%2C%22central%20portion%20of%20breast%22%2C%22lower-inner%20quadrant%20of%20breast%22%2C%22lower-outer%20quadrant%20of%20breast%22%2C%22nipple%22%2C%22overlapping%20lesion%20of%20breast%22%2C%22upper-inner%20quadrant%20of%20breast%22%2C%22upper-outer%20quadrant%20of%20breast%22%5D%7D%2C%22op%22%3A%22in%22%7D%2C%7B%22content%22%3A%7B%22field%22%3A%22cases.primary_site%22%2C%22value%22%3A%5B%22breast%22%5D%7D%2C%22op%22%3A%22in%22%7D%2C%7B%22op%22%3A%22in%22%2C%22content%22%3A%7B%22field%22%3A%22cases.project.program.name%22%2C%22value%22%3A%5B%22TCGA%22%5D%7D%7D%5D%7D&searchTableTab=genes), based on 1077 cases.

# Obtain gene-stratified survival

```{r dogs}
gstrat = function(sym="TTN", mutlist, survdat) {
 stopifnot(inherits(survdat, "Surv"))
 stopifnot(length(survdat) == length(mutlist))
 hassym <- unlist(list(sym) %in% mutlist)
 plot(survfit(survdat~hassym), main=sym, lty=1:2, xlab="Years")
}
gstrat("TTN", mutsyms, osurv)
gstrat("TP53", mutsyms, osurv)
```

# Frequencies

```{r dofr}
commut = tail(sort(table(unlist(mutsyms))), 30)
commut
```

```{r lkhist}
hist(log(sapply(mutsyms,length)), main="Log mutation count per tumor")
```
 
# Combinations of mutations

```{r lkco}
hasmut = function(sym="TP53", mutlist) {
 sapply(mutlist, function(x) sym %in% x)
}
table( hasmut("TP53", mutsyms), hasmut("PIK3CA", mutsyms) )
```

Assemble pairs
```{r lkmpairs}
common_pairs = combn(names(commut),2)
common_pairs[,1:4]
```

# Greedy search for deleterious pairs

```{r lkpairs}
indicate_pair = function(sym1, sym2, mutlist)
 hasmut(sym1, mutlist) & hasmut(sym2, mutlist)
chk = apply(common_pairs,2,function(z) indicate_pair(z[1], z[2], mutsyms))
chkp.inds = which(apply(chk,2,sum)>20) # disallow very rare combos
dim(chk)
dim(chk[,chkp.inds])
chisqs = apply(chk[,chkp.inds],2,function(z)survdiff(osurv~z)$chisq)
hist(chisqs)
cpr = common_pairs[,chkp.inds][, which(chisqs>5)]
plot(survfit(osurv~chk[,chkp.inds[which(chisqs>5)]]), lty=1:2, main=paste(cpr, collapse=" & " ))
```

# Using domain classification of mutations

```{r lkdom, cache=TRUE}
dom = assay(bmut.surv[[1L]], "domain_WU")
length(grep("Znf", na.omit(as.character(dom)))) # frequently noted
system.time(
    mutdoms <- bplapply(1:ncol(dom), function(x) as.character(na.omit(dom[,x])))
)
```

Isolate individuals with mutations annotated to Znf domain

```{r noz}
noz = sapply(mutdoms, function(x) length(grep("Znf", x))==0)
table(noz)
survdiff(osurv~noz)
```

SH3

```{r nosh}
nosh3 = sapply(mutdoms, function(x) length(grep("SH3", x))==0)
survdiff(osurv~nosh3)
```

